---
title: "Assignment 1"
author: "Conor Mangan"
output: html_document
---

## Introduction

This assignment demonstrates the use of R for importing,
describing, transforming, and visualizing data. The dataset
used in this assignment is the Palmer Penguins dataset.

The dataset is available from the Palmer Station Long-Term
Ecological Research program.

## Importing the Data

```{r}
library(palmerpenguins)

data(penguins)

head(penguins)
## Descriptive Statistics
```
summary(penguins[, c("bill_length_mm",
                     "bill_depth_mm",
                     "flipper_length_mm",
                     "body_mass_g")])
### Categorical Variables
```{r}
table(penguins$species)

table(penguins$island)
```


```{r}
table(penguins$sex)
```
penguins$body_mass_kg <- penguins$body_mass_g / 1000

head(penguins[, c("body_mass_g", "body_mass_kg")])

## Quantitative Variable Plot

The following histogram displays the distribution of penguin body mass.

```{r}
hist(penguins$body_mass_g,
     main = "Distribution of Penguin Body Mass",
     xlab = "Body Mass (grams)")
```

## Scatterplot

The following scatterplot examines the relationship between flipper length and body mass.

```{r}
plot(penguins$flipper_length_mm,
     penguins$body_mass_g,
     main = "Flipper Length vs. Body Mass",
     xlab = "Flipper Length (mm)",
     ylab = "Body Mass (g)")
```
```{r}

```
