---
title: "Assignment 2: k-NN Classification"
author: "Conor Mangan"
date: "2026-09-30"
output: html_document
---

# Introduction

The purpose of this analysis is to use k-nearest neighbors (k-NN)
classification to predict whether a Universal Bank customer will accept
a personal loan offer. The dataset contains information on 5,000
customers, including demographic characteristics, their relationship
with the bank, and their response to a previous personal loan campaign.
The response variable is Personal Loan, where a value of 1 indicates
that the customer accepted the personal loan offer and a value of 0
indicates that the customer did not accept the offer.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(class)
library(dplyr)

set.seed(1)
```

# Load and Examine the Data

The Universal Bank dataset is imported and examined below.

```{r}
bank <- read.csv("UniversalBank.csv")

head(bank)
dim(bank)
str(bank)
table(bank$Personal.Loan)
```

The dataset contains 5,000 customers. Of these customers, 480 accepted
the previous personal loan offer, representing 9.6% of the observations.

# Data Preparation

The ID and ZIP Code variables are not used as predictors. Education has
three categories, so it is transformed into three dummy variables before
performing k-NN classification.

```{r}
bank$Education_1 <- ifelse(bank$Education == 1, 1, 0)
bank$Education_2 <- ifelse(bank$Education == 2, 1, 0)
bank$Education_3 <- ifelse(bank$Education == 3, 1, 0)

bank_knn <- bank %>%
  select(
    Age,
    Experience,
    Income,
    Family,
    CCAvg,
    Education_1,
    Education_2,
    Education_3,
    Mortgage,
    Securities.Account,
    CD.Account,
    Online,
    CreditCard,
    Personal.Loan
  )

head(bank_knn)
```

# Question 1

The data are first partitioned into 60% training and 40% validation
sets. A random seed is specified so that the partition can be
reproduced.

```{r}
set.seed(1)

train_index <- createDataPartition(
  bank_knn$Personal.Loan,
  p = 0.60,
  list = FALSE
)

train_data <- bank_knn[train_index, ]
valid_data <- bank_knn[-train_index, ]

nrow(train_data)
nrow(valid_data)
```

The predictor variables and response variable are separated.

```{r}
train_x <- train_data %>%
  select(-Personal.Loan)

valid_x <- valid_data %>%
  select(-Personal.Loan)

train_y <- train_data$Personal.Loan
valid_y <- valid_data$Personal.Loan
```

Because k-NN relies on distances between observations, the predictor
variables are standardized. The standardization parameters are
estimated using only the training data and are then applied to the
validation data.

```{r}
preprocess <- preProcess(
  train_x,
  method = c("center", "scale")
)

train_x_norm <- predict(preprocess, train_x)
valid_x_norm <- predict(preprocess, valid_x)
```

The customer specified in the assignment is created below.

```{r}
new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education_1 = 0,
  Education_2 = 1,
  Education_3 = 0,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1
)

new_customer_norm <- predict(preprocess, new_customer)
```

A k-NN model with k = 1 is used to classify the customer.

```{r}
set.seed(1)

prediction_k1 <- knn(
  train = train_x_norm,
  test = new_customer_norm,
  cl = train_y,
  k = 1
)

prediction_k1
```

### Question 1 Interpretation

Using k = 1, the customer was classified as **`r prediction_k1`**.
A classification of 1 indicates that the customer is predicted to
accept the personal loan offer, while a classification of 0 indicates
that the customer is predicted not to accept the personal loan offer.

# Question 2

To determine an appropriate value of k, values from 1 through 30 are
evaluated using the validation data. Validation accuracy is calculated
for each value.

```{r}
k_values <- 1:30
accuracy_values <- numeric(length(k_values))

for (i in seq_along(k_values)) {

  pred <- knn(
    train = train_x_norm,
    test = valid_x_norm,
    cl = train_y,
    k = k_values[i]
  )

  accuracy_values[i] <- mean(pred == valid_y)
}

results <- data.frame(
  k = k_values,
  Accuracy = accuracy_values
)

results
```

The value of k with the highest validation accuracy is identified.

```{r}
best_k <- results$k[which.max(results$Accuracy)]
best_accuracy <- max(results$Accuracy)

best_k
best_accuracy
```

Validation accuracy across the different values of k is displayed
graphically.

```{r}
plot(
  results$k,
  results$Accuracy,
  type = "b",
  xlab = "Number of Neighbors (k)",
  ylab = "Validation Accuracy",
  main = "Validation Accuracy for Different Values of k"
)
```

### Question 2 Interpretation

Very small values of k can lead to overfitting because predictions
depend on only a small number of neighboring observations. Conversely,
very large values of k can ignore useful local patterns in the
predictor information. Based on the validation results, the selected
value is **k = `r best_k`**, which produced a validation accuracy of
**`r round(best_accuracy * 100, 2)`%**.

# Question 3

The selected value of k is now used to classify the observations in the
validation set. The resulting confusion matrix is shown below.

```{r}
best_pred_valid <- knn(
  train = train_x_norm,
  test = valid_x_norm,
  cl = train_y,
  k = best_k
)

cm_validation <- confusionMatrix(
  factor(best_pred_valid, levels = c(0, 1)),
  factor(valid_y, levels = c(0, 1)),
  positive = "1"
)

cm_validation
```

### Question 3 Interpretation

The confusion matrix compares the predicted personal loan classifications
with the actual outcomes in the validation data. Correct predictions
appear along the diagonal of the matrix, while false positives and
false negatives appear off the diagonal. Because only a relatively
small percentage of customers accepted the previous personal loan
offer, both overall accuracy and the model's ability to identify
customers in the acceptance class should be considered when evaluating
performance.

# Question 4

The customer from Question 1 is classified again using the selected
value of k.

```{r}
prediction_best_k <- knn(
  train = train_x_norm,
  test = new_customer_norm,
  cl = train_y,
  k = best_k
)

prediction_best_k
```

### Question 4 Interpretation

Using **k = `r best_k`**, the customer is classified as
**`r prediction_best_k`**. A classification of 1 indicates predicted
acceptance of the personal loan offer, while a classification of 0
indicates predicted non-acceptance.

# Question 5

The data are now repartitioned into 50% training, 30% validation, and
20% test sets. The same value of k selected previously is used for this
analysis.

```{r}
set.seed(1)

train_index2 <- createDataPartition(
  bank_knn$Personal.Loan,
  p = 0.50,
  list = FALSE
)

train2 <- bank_knn[train_index2, ]
remaining <- bank_knn[-train_index2, ]

valid_index2 <- createDataPartition(
  remaining$Personal.Loan,
  p = 0.60,
  list = FALSE
)

valid2 <- remaining[valid_index2, ]
test2 <- remaining[-valid_index2, ]

nrow(train2)
nrow(valid2)
nrow(test2)
```

The predictor and response variables are separated for each partition.

```{r}
train2_x <- train2 %>%
  select(-Personal.Loan)

valid2_x <- valid2 %>%
  select(-Personal.Loan)

test2_x <- test2 %>%
  select(-Personal.Loan)

train2_y <- train2$Personal.Loan
valid2_y <- valid2$Personal.Loan
test2_y <- test2$Personal.Loan
```

The predictors are standardized using the training data.

```{r}
preprocess2 <- preProcess(
  train2_x,
  method = c("center", "scale")
)

train2_norm <- predict(preprocess2, train2_x)
valid2_norm <- predict(preprocess2, valid2_x)
test2_norm <- predict(preprocess2, test2_x)
```

Predictions are generated for the training, validation, and test sets
using the previously selected value of k.

```{r}
pred_train2 <- knn(
  train = train2_norm,
  test = train2_norm,
  cl = train2_y,
  k = best_k
)

pred_valid2 <- knn(
  train = train2_norm,
  test = valid2_norm,
  cl = train2_y,
  k = best_k
)

pred_test2 <- knn(
  train = train2_norm,
  test = test2_norm,
  cl = train2_y,
  k = best_k
)
```

## Training Confusion Matrix

```{r}
cm_train <- confusionMatrix(
  factor(pred_train2, levels = c(0, 1)),
  factor(train2_y, levels = c(0, 1)),
  positive = "1"
)

cm_train
```

## Validation Confusion Matrix

```{r}
cm_valid2 <- confusionMatrix(
  factor(pred_valid2, levels = c(0, 1)),
  factor(valid2_y, levels = c(0, 1)),
  positive = "1"
)

cm_valid2
```

## Test Confusion Matrix

```{r}
cm_test <- confusionMatrix(
  factor(pred_test2, levels = c(0, 1)),
  factor(test2_y, levels = c(0, 1)),
  positive = "1"
)

cm_test
```

The overall accuracy for each partition is extracted below to make the
results easier to compare.

```{r}
train_accuracy <- cm_train$overall["Accuracy"]
valid_accuracy <- cm_valid2$overall["Accuracy"]
test_accuracy <- cm_test$overall["Accuracy"]

accuracy_comparison <- data.frame(
  Data_Set = c("Training", "Validation", "Test"),
  Accuracy = c(
    train_accuracy,
    valid_accuracy,
    test_accuracy
  )
)

accuracy_comparison
```

### Question 5 Interpretation

Using **k = `r best_k`**, the training set produced an accuracy of
**`r round(train_accuracy * 100, 2)`%**, the validation set produced an
accuracy of **`r round(valid_accuracy * 100, 2)`%**, and the test set
produced an accuracy of **`r round(test_accuracy * 100, 2)`%**.

Differences among the three confusion matrices are expected because
each partition contains different observations. The training
observations are directly involved in constructing the k-NN
classifier, whereas the validation and test observations provide an
evaluation of the model on observations that were not used to construct
the classifier. Similar performance between the validation and test
sets would indicate that the model generalizes reasonably well to new
customers.

The class imbalance should also be considered when interpreting these
results. Only 480 of the 5,000 customers, or 9.6%, accepted the
personal loan in the previous campaign. Therefore, overall accuracy
alone does not provide a complete picture of model performance, and
the false positives, false negatives, sensitivity, and specificity
shown in the confusion matrices are also useful for evaluating the
classifier.

# Conclusion

This analysis applied k-nearest neighbors classification to the
Universal Bank customer data to predict personal loan acceptance.
The model was initially evaluated using a 60% training and 40%
validation partition, and different values of k were compared to select
an appropriate number of neighbors. The selected model was then used
to classify the specified customer and was further evaluated using
training, validation, and test sets. The confusion matrices provide
information about both correct and incorrect classifications and help
evaluate how well the k-NN model may generalize to new customers.